Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-17 mr-review.Rmd
updated timeline inputs

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Two-sample 
Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian]. They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects. Importantly, the SNP instruments are measured in both sets of subjects.

@@

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
are ranked by the strength of their marginal association with the exposure 
variable. The LD clumping procedure then outputs high ranking SNPs that are not in high LD with each other.

@@

@@

The estimated marginal effects for the $p$ SNP instruments on the exposure are denoted by $\hat \beta_X$, while 
$\hat\beta_Y$ represents the estimated effects of the same SNPs on the outcome. The $p$ by $p$ correlation matrices for 
the SNP instruments are written as $\Sigma_1$ and $\Sigma_2$, respectively. $\Sigma_1$ and $\Sigma_2$ can be estimated with 1000 Genomes 
Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry 
[@the_1000_genomes_project_consortium_1000_2012]. The $p$-vector error terms, $e_x$ and $e_y$, follow multivariate normal 
distributions with mean zero and variances $\Sigma_1\frac{\sigma_X^2}{n_1 - 1}$ 
and $\Sigma_2\frac{\sigma_Y^2}{n_2 - 1}$, 
respectively. @yuan_likelihood_2021 then constructs a Gibbs sampler to make likelihood-based inferences [@geman1984stochastic]. To facilitate computations, the investigators make several assumptions about the collection of SNPs:

>>>>

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Two-sample 
Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian]. They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.

@@

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
are ranked by the strength of their marginal association with the exposure 
variable.

@@

```{r timeline, fig.cap="Timeline of two-sample GWAS summary statistics MR methods"}
readxl::read_xlsx("data/method-year.xlsx", sheet = "inputs")
```

@@

The estimated marginal effects for the $p$ SNP instruments on the exposure are denoted by $\hat \beta_X$, while 
$\hat\beta_Y$ represents the estimated effects of the same SNPs on the outcome. The $p$ by $p$ SNP instrument correlation matrices for the exposure and outcome 
are written as $\Sigma_1$ and $\Sigma_2$, respectively. $\Sigma_1$ and $\Sigma_2$ can be estimated with 1000 Genomes 
Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry 
[@the_1000_genomes_project_consortium_1000_2012]. The $p$-vector error terms, $e_x$ and $e_y$, follow multivariate normal 
distributions with mean zero and variances $\Sigma_1\frac{\sigma_X^2}{n_1 - 1}$ 
and $\Sigma_2\frac{\sigma_Y^2}{n_2 - 1}$, 
respectively. @yuan_likelihood_2021 then constructs a Gibbs sampler to make likelihood-based inferences [@geman1984stochastic]. To facilitate computations, the investigators make several assumptions about the collection of SNPs:
